Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-05 analysis/paper/paper.qmd
updates

library(data.table)
library(NISTunits)
library(circular)
library(ggpubr)
library(ggalt)
library(ggplot2)
library(stringr)
library(ggtern)
library(dplyr)
library(here)
source(here('analysis/paper/scripts/McPherron_2018.r'))
source(here('analysis/paper/scripts/get_coordinates.r'))
source(here('analysis/paper/scripts/vector_stats.r'))
```

@@

At present, the debate remains stalled, primarily due to the lack of clear typological criteria for distinguishing spheroids intentionally shaped according to conceptual templates from those that emerged as by-products of flaking or percussive activities. Regional studies continue to rely heavily on individual researchers’ observational judgment and experiential knowledge. Furthermore, current typological approaches offer limited capacity for cross-regional comparison, making it difficult for researchers to verify interpretations derived from other sites. At the same time, conclusions drawn from their own analyses are rarely subject to external validation or replication.

To address these challenges, we developed a set of quantitative methods combining morphological and technological analyses to investigate spheroid production strategies. While previous morgphlogical measures have focused on the shape of spheroids close to perfect shphere, contributing to discussions of spheroid standardization, they cannot reliably reflect production strategies, as highly spherical forms may result from diverse production goals. We therefore adopted the novel spherical harmonics, a landmark-free method recently introduced to archaeology, which offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks. We have creatively modified the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons—this is the first attempt of its kind in the field of archaeology. Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production strategies. <!-- BM: highlight open source aspect -->

@@

The Qianshangying site complex consists of four localities from QSY-A to QSY-D, discovered in 2015. It is among the region’s most artifact-rich archaeological site. Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode <!-- BM: citation needed -->. The stratigraphic context of Qianshangying indicates a short-term occupation resulting from a single hominin activity event. Analyses of artefact orientations and size distributions indicates that the assemblage appears to be largely _in situ_, exhibiting minimal disturbance after deposition  <!-- BM: cite your excavation report, unpublished report -->. ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka <!-- BM: cite your excavation report, unpublished report -->.

@@

Raw materials at Qianshangying. were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others. Lava dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively). Spheroids show no strong raw material preference. Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert. However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores. <!-- BM: consider a multipanel figure, bar chart for raw material, bar chart for basic typology, consider four plots we can combine to give a basic overview of the assemblage to to prepare the reader for the more specialised data coming later in the paper -->

@@

long_data <- data %>%
  select(core_length, core_width, core_thickness,
         spheroid_length, spheroid_width, spheroid_thickness,
         multi_poly_length, multi_poly_width, multi_poly_thickness) %>%
  pivot_longer(
    everything(),
    names_to = "type_metric",
    values_to = "value"
  ) %>%
  extract(type_metric, into = c("artifact", "metric"),
          regex = "^(core|spheroid|multi_poly)_(length|width|thickness)") %>%
  filter(!is.na(value)) %>%
  mutate(
    artifact = recode(artifact,
                      "core" = "Core",
                      "spheroid" = "Spheroid",
                      "multi_poly" = "Multi-poly Core"),
    metric = factor(metric, levels = c("length", "width", "thickness"))
  )

# 画图
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Type"
  ) +
  scale_fill_manual(values = c("Core" = "#66c2a5", 
                               "Spheroid" = "orange", 
                               # BM: what is 'Multi-poly'? we don't have this word anywhere else in the text
                               "Multifacial Core" = "lightgray")) +
  theme_minimal(base_size = 14)

@@

This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts. @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant. @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items. Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250). Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced. Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM. We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.

@@

To reduce the dimensionality and visualize the structure of shape variation, we applied a Uniform Manifold Approximation and Projection (UMAP) to the normalized power spectrum. The UMAP reveals distinct clusters in the first two dimensions of the shape space. These clusters correspond to reference models, sphere, rounded cube, ellipsoid, box and disc-shaped core (@fig-UMAP-SPHARM-clusters). This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features. Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples). However, some Multifacial cores and Polyhedron also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids. <!-- BM: this is an excellent analysis of these results! I think we might be able to combine them into one figure, two plots -->

>>>>

library(data.table)
library(NISTunits)
library(circular)
library(ggpubr)
library(ggalt)
library(ggplot2)
library(stringr)
library(ggtern)
library(dplyr)
library(here)
library(tidyverse)
source(here('analysis/paper/scripts/McPherron_2018.r'))
source(here('analysis/paper/scripts/get_coordinates.r'))
source(here('analysis/paper/scripts/vector_stats.r'))
```

@@

At present, the debate remains stalled, primarily due to the lack of clear typological criteria for distinguishing spheroids intentionally shaped according to conceptual templates from those that emerged as by-products of flaking or percussive activities. Thus, regional studies continue to rely heavily on individual researchers’ observational judgment and experiential knowledge. Furthermore, current typological approaches offer limited capacity for cross-regional comparison, making it difficult for researchers to verify interpretations derived from other sites. At the same time, conclusions drawn from their own analyses are rarely subject to external validation or replication.

To address these challenges, we developed a set of quantitative methods combining morphological and technological analyses to investigate spheroid production strategies. While previous morphlogical measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they cannot reliably reflect production strategies, as highly spherical forms may result from diverse production goals. We therefore adopted the novel spherical harmonics, a landmark-free method recently introduced to archaeology, which offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks. We have creatively modified the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons—this is the first attempt of its kind in the field of archaeology. Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production strategies. All of these methods are open-source to enable replication and support a unified quantitative research framework.

@@

df <- tibble(
  material = c("Lava", "Dolomite", "Chert", "Other"),
  spheroids = c(9, 2, 0, 1),
  cores = c(62, 21, 4, 7),
  total = c(637, 209, 53, 56),
)

df_long <- df %>%
  pivot_longer(cols = -material, names_to = "category", values_to = "count") %>%
  group_by(category) %>%
  mutate(prop = count / sum(count) * 100) %>%
  ungroup() %>%
  mutate(
    category = factor(category, levels = c("spheroids", "cores", "total")),
    material = factor(material, levels = c("Lava", "Dolomite", "Chert", "Other"))
  )

ggplot(df_long, aes(x = material, y = prop, fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    title = "Raw Material Proportions in Spheroids, Cores, and Total",
    x = "Raw Material",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  scale_fill_manual(values = c(
    "spheroids" = "orange", 
    "cores" = "#66c2a5", 
    "total" = "lightgray"
  )) +
  theme_minimal() +
  theme(
    legend.position = c(0.95, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black")  
  )

#citation
df_long <- df_long %>%
  mutate(text = paste0("n = ", count, ", ", round(prop, 1), "%"))
#print(df_long)
lava_total_text <- df_long %>% filter(material == "Lava", category == "total") %>% pull(text)
dolomite_total_text <- df_long %>% filter(material == "Dolomite", category == "total") %>% pull(text)
chert_total_text <- df_long %>% filter(material == "Chert", category == "total") %>% pull(text)
other_total_text <- df_long %>% filter(material == "Other", category == "total") %>% pull(text)

lava_spheroids_text <- df_long %>% filter(material == "Lava", category == "spheroids") %>% pull(text)
dolomite_spheroids_text <- df_long %>% filter(material == "Dolomite", category == "spheroids") %>% pull(text)
chert_spheroids_text <- df_long %>% filter(material == "Chert", category == "spheroids") %>% pull(text)
other_spheroids_text <- df_long %>% filter(material == "Other", category == "spheroids") %>% pull(text)
```

The Qianshangying site complex consists of four localities from QSY-A to QSY-D, discovered in 2015. It is among the region’s most artifact-rich archaeological site. Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode <!-- BM: citation needed -->. The stratigraphic context of Qianshangying indicates a short-term occupation resulting from a single hominin activity event. Analyses of artefact orientations and size distributions indicates that the assemblage appears to be largely *in situ*, exhibiting minimal disturbance after deposition <!-- BM: cite your excavation report, unpublished report -->. ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka <!-- BM: cite your excavation report, unpublished report -->.

@@

Raw materials at Qianshangying were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others. The overall assemblage is dominated by lava ( `r lava_total_text` ), followed by siliceous dolomite ( `r dolomite_total_text` ), with smaller proportions of chert ( `r chert_total_text` ) and other materials ( `r other_total_text` ). The composition of cores closely mirrors this distribution, comprising `r round(df_long$prop[df_long$material == "Lava" & df_long$category == "cores"],1)`% lava (n = `r df_long$count[df_long$material == "Lava" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Dolomite" & df_long$category == "cores"],1)`% siliceous dolomite (n = `r df_long$count[df_long$material == "Dolomite" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Chert" & df_long$category == "cores"],1)`% chert (n = `r df_long$count[df_long$material == "Chert" & df_long$category == "cores"]`), and `r round(df_long$prop[df_long$material == "Other" & df_long$category == "cores"],1)`% other (n = `r df_long$count[df_long$material == "Other" & df_long$category == "cores"]`). In contrast, spheroids exhibit a marked preference for lava (`r lava_spheroids_text`), with reduced use of siliceous dolomite ( `r dolomite_spheroids_text` ) and no use of chert. Compared to both the total assemblage and the cores, spheroids show a stronger reliance on lava and a diminished use of more fragile materials, suggesting a selective strategy favoring raw materials with more stable and homogeneous physical properties.

@@

long_data <- data %>%
  select(core_length, core_width, core_thickness,
         spheroid_length, spheroid_width, spheroid_thickness,
         multi_poly_length, multi_poly_width, multi_poly_thickness) %>%
  pivot_longer(
    everything(),
    names_to = "type_metric",
    values_to = "value"
  ) %>%
  extract(type_metric, into = c("artifact", "metric"),
          regex = "^(core|spheroid|multi_poly)_(length|width|thickness)") %>%
  filter(!is.na(value)) %>%
  mutate(
    artifact = recode(artifact,
                      "core" = "Core",
                      "spheroid" = "Spheroid",
                      "multi_poly" = "Polys and Multi cores"),
    metric = factor(metric, levels = c("length", "width", "thickness"))
  )

# 画图
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Type"
  ) +
  scale_fill_manual(values = c(
    "Core" = "#66c2a5", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "lightgray"
  )) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = c(0.95, 0.95),       
    legend.justification = c("right", "top"),  
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA), 
    legend.box.background = element_rect(color = "black")  
  )

@@

This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts. @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant. @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items. Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.\@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250). Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced. Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM. We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.

@@

To reduce the dimensionality and visualize the structure of shape variation, we applied a Uniform Manifold Approximation and Projection (UMAP) to the normalized power spectrum. The UMAP reveals distinct clusters in the first two dimensions of the shape space. These clusters correspond to reference models, sphere, rounded cube, ellipsoid, box and disc-shaped core (@fig-UMAP-SPHARM-clusters). This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features. Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples). However, some Multifacial cores and Polyhedron also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.
